### 2017-01-20 thesis/methods/methods.tex
Finished obesity metagene direction subsection

In order to assess whether a certain set of genes are up or down regulated by a sample, some sort of score based on these genes must be calculated.
\Gls{svd} is a mathematical method that splits a matrix into several matrices that represent the original matrix \citep{Golub1970}.
With \gls{svd}, a matrix $X$ of size $n$ genes by $m$ arrays (or samples) can be represented as:
\begin{equation}
	\label{eq:svd}
	X = UDV^T
\end{equation}
where the columns of $U$ and $V$ contain the left and right singular vectors of $X$ respectively, and $D$ contains the singular values of $X$.

@@

\section{Gatza pathway metagene direction}
\label{sec:pathway_metagene_direction}

All 18 pathway signatures from \citet{Gatza2010a} study were used to create both the metagenes and transformation matrices in the Gatza data set.
Each of the 18 pathway metagenes were compared with the gene expression of the gene that represents the pathway to check whether the metagenes were in the correct direction.
% TODO: check if it was the AKT1 or AKT2 gene for AKT pathway
For example, the AKT pathway metagene was compared with the AKT1 gene expression.

The pathway metagene value was first compared visually with its pathway gene in a heatmap.
If the direction of the pathway metagene was correct (for example, high pathway metagene value corresponded with high expression of the gene representing that pathway), the metagene values were not changed, but if it was incorrect, the metagene values were flipped for that pathway.

Spearman correlation of the pathway metagene and the expression of the pathway gene was also used to distinguish whether the pathway metagene was in the right direction.
The direction of the metagene was considered correct when the correlation was positive and incorrect when the correlation was negative.

The clustering of the pathway metagenes were taken into account to decide on the final direction of the metagene.
The direction of the pathway metagenes were modified until the resulting pathway metagenes were similarly clustered together as in the results from \citet{Gatza2010a} paper.

>>>>

\subsection{\Gls{svd}}
\label{sub:svd}

In order to ertain set of genes are up or down regulated by a sample, some sort of score based on these genes must be calculated.
\Gls{svd} is a mathematical method that splits a matrix into several matrices that represent the original matrix \citep{Golub1970}.
With \gls{svd}, a matrix $X$ of size $n$ genes by $m$ arrays (or samples) can be represented as:
\begin{equation}
	\label{eq:svd}
	X = UDV^T
\end{equation}
where the columns of $U$ and $V$ contain the left and right singular vectors of $X$ respectively, and $D$ contains the singular values of $X$.

@@

\subsection{Metagene direction}
\label{sub:metagene_direction}

One thing to consider when analysing data with metagenes is the direction of the metagene.
When \gls{svd} creates the metagene from the given data, it does not consider the direction of the signal, but only the magnitude of the signal in the data.
Therefore, the metagene created by \gls{svd} can be either positively or negatively correlated to the phenotype that the set of genes reflect.
For example, a metagene created from a set of genes that are associated with \gls{er} status may have higher metagene scores in the \gls{er} negative samples and low metagene scores in the \gls{er} positive samples.
In this case, the generated metagene must be flipped in order to reflect the phenotype with the metagene scores (low metagene score with low phenotype and high metagene score with high phenotype).

In this project, two types of metagenes were created: obesity associated and pathway associated metagenes.
For obesity associated metagenes, the direction of the metagenes were checked with the obesity status and \gls{bmi} value of the samples from the data in which the metagene was derived from.
Checking the direction of the pathway metagenes from \citet{Gatza2010a} study is not as simple as the obesity associated metagenes, as the experimental design that identify which samples were treated to generate the pathway signature was not available.
Due to this, it was not possible to validate whether the generated pathway metagene was in the right direction.

As an alternative, the pathway metagenes were visually compared with the gene that represent the pathway using a heatmat, as the expression of this gene would most likely be affected the most by the pathway signature.
For example, the Akt pathway metagene was compared with the AKT1 gene expression, and if the metagene was not in the same direction as the AKT1 gene expression, the metagene was flipped.
\cref{tab:metagene_direction} summarises the genes used to check the direction of the pathway metagenes.
\begin{table}[htb]
	\caption{18 pathways from \citet{Gatza2010a} and its respective genes used to check the direction of the pathway metagene}
	\label{tab:metagene_direction}
	\begin{center}
		\begin{tabular}{lc}
			Pathway & Representative gene\\
			\hline
			\rule{0pt}{2.25ex}Akt & AKT1\\
			\gls{bcat}            & CTNNB1\\
			E2F1                  & E2F1\\
			\gls{egfr}            & EGFR\\
			\gls{er}              & ESR1\\
			\gls{her2}            & ERBB2\\
			\gls{ifna}            & IFNA1\\
			\gls{ifny}            & IFNG\\
			Myc                   & MYC\\
			p53                   & TP53\\
			p63                   & TP63\\
			\gls{pi3k}            & PIK3CA\\
			\gls{pr}              & PGR\\
			Ras                   & HRAS\\
			\gls{stat3}           & STAT3\\
			Src                   & SRC\\
			\gls{tgfb}            & TGFB1\\
			\gls{tnfa}            & TNF\\
			\hline
			\hline
		\end{tabular}
	\end{center}
\end{table}
In addition to this, the correlation of the pathway metagene with its repective gene was checked to further aid the decision on whether the metagene is going in the correct direction.
This was due to the fact that some of the metagenes were visually difficult to determine whether it was in the right direction.
Therefore, the direction as well as the magnitude of the correlation value was also taken into account in making the decision whether to flip the metagene.
Lastly, further on top of this, the clustering of the pathway metagenes with the other pathway metagenes were also considered.
It was known from \citet{Gatza2010a} study that some pathways clustered and grouped together with specific pathways, so the metagenes were flipped so that the grouping of the pathways matched the results from their study.
